---
title: "Week 4 - Homework"
author: "STAT 420, Summer 2018, Unger"
date: ''
output:
  html_document: 
    toc: yes
  pdf_document: default
urlcolor: cyan
---

***Name*** : Hussein Elmessilhy

***NetID***: hae2


## Exercise 1 (Using `lm`)

For this exercise we will use the data stored in [`nutrition-2018.csv`](nutrition-2018.csv). It contains the nutritional values per serving size for a large variety of foods as calculated by the USDA in 2018. It is a cleaned version totaling 5956 observations and is current as of April 2018.

The variables in the dataset are:

- `ID` 
- `Desc` - short description of food
- `Water` - in grams
- `Calories` 
- `Protein` - in grams
- `Fat` - in grams
- `Carbs` - carbohydrates, in grams
- `Fiber` - in grams
- `Sugar` - in grams
- `Calcium` - in milligrams
- `Potassium` - in milligrams
- `Sodium` - in milligrams
- `VitaminC` - vitamin C, in milligrams
- `Chol` - cholesterol, in milligrams
- `Portion` - description of standard serving size used in analysis

**(a)** Fit the following multiple linear regression model in `R`. Use `Calories` as the response and `Fat`, `Sugar`, and `Sodium` as predictors.

\[
Y_i = \beta_0 + \beta_1 x_{i1} + \beta_2 x_{i2} + \beta_3 x_{i3} + \epsilon_i.
\]

Here,

- $Y_i$ is `Calories`.
- $x_{i1}$ is `Fat`.
- $x_{i2}$ is `Sugar`.
- $x_{i3}$ is `Sodium`.

Use an $F$-test to test the significance of the regression. Report the following:
 
- The null and alternative hypotheses
- The value of the test statistic
- The p-value of the test
- A statistical decision at $\alpha = 0.01$
- A conclusion in the context of the problem

When reporting these, you should explicitly state them in your document, not assume that a reader will find and interpret them from a large block of `R` output.

***Answer***

```{r}
setwd(".\\")
nutrition_data <- read.csv("nutrition-2018.csv", header = TRUE, sep = ",")
nutrition_model <- lm(Calories ~ Fat + Sugar + Sodium , nutrition_data)
nutrition_model_summary <- summary(nutrition_model)

```

- The null and alternative hypotheses
$H_0: \beta_1 = \beta_2 = \beta_3 = 0$

$H_1: \texttt{At least one of } \beta_j, j = 1,2,3 \neq 0$

- The value of the test statistic

`r nutrition_model_summary$fstatistic[1]`

- The p-value of the test

2.2e-16

- A statistical decision at $\alpha = 0.01$

As the p-value is less than alpha, statistical decision is to reject the null hypotheses

- A conclusion in the context of the problem

Conclusion is that the regression is significant, which means that there's a linear relationship between the calories in food anda amount of fat, sugar, and sodium it has

**(b)** Output only the estimated regression coefficients. Interpret all $\hat{\beta}_j$ coefficients in the context of the problem.

***Answer***

$\hat{\beta}_1$ is `r summary(nutrition_model)$coefficients[2,1]`, it corresponds to the average change in calories for an increase of one gram of fats for a food of certain amount of sugar and sodium. 

$\hat{\beta}_2$ is `r summary(nutrition_model)$coefficients[3,1]`, it corresponds to the average change in calories for an increase of one gram of sugar for a food of certain amount of fat and sodium. 

$\hat{\beta}_3$ is `r summary(nutrition_model)$coefficients[4,1]`, it corresponds to the average change in calories for an increase of one milligram of sodium for a food of certain amount of sugar and fat 


**(c)** Use your model to predict the number of `Calories` in a Big Mac. According to [McDonald's publicized nutrition facts](https://www.mcdonalds.com/us/en-us/about-our-food/nutrition-calculator.html), the Big Mac contains 28g of fat, 9g of sugar, and 950mg of sodium.

***Answer***
```{r}
predict(nutrition_model, newdata = data.frame(Fat = 28, Sugar = 9, Sodium = 950))
```

**(d)** Calculate the standard deviation, $s_y$, for the observed values in the Calories variable. Report the value of $s_e$ from your multiple regression model. Interpret both estimates in the context of this problem.


***Answer***

The two requested numbers $s_y$ and $s_e$ respectively are:
`r c(sd(nutrition_data$Calories), summary(nutrition_model)$sigma)`

While $s_y$ value represents the amount of dispersion in the calories around their mean  regardless of the model, $s_e$ which is the residual standard error, denotes that our estimates of calories by the regression model are "typically" off by the second amount.


**(e)** Report the value of $R^2$ for the model. Interpret its meaning in the context of the problem.

***Answer***
$R^2$ is `r summary(nutrition_model)$r.squared`
Interpretation : this proportion of the observed variability in amount of calories is explained by the linear relationship with fats, sugar, and sodium as defined by this model.

**(f)** Calculate a 95% confidence interval for $\beta_2$. Give an interpretation of the interval in the context of the problem.

***Answer***
95% confident interval for $\beta_2$ is :
```{r}
c(confint(nutrition_model, level = 0.95)[3,1], confint(nutrition_model, level = 0.95)[3,2])
```
And it means that we are 95% that the mean change in calories for an increase of one gram of sugar given a certain amount of fat and sodium is between these two values.

**(g)** Calculate a 99% confidence interval for $\beta_0$. Give an interpretation of the interval in the context of the problem.

***Answer***
99% confident interval (lower, upper) for $\beta_2$ is :
```{r}
c(confint(nutrition_model, level = 0.99)[1,1], confint(nutrition_model, level = 0.99)[1,2])
```
And it means that we are 99% that number of calories for food with zero fat, sugar, and sodium is between these two values.

**(h)** Use a 90% confidence interval to estimate the mean Calorie content of a food with 24g of fat, 0g of sugar, and 350mg of sodium, which is true of a large order of McDonald's french fries. Interpret the interval in context.
 
 ***Answer***
```{r}
predict(nutrition_model, level = 0.9, interval = 'confidence', newdata = data.frame(Fat = 24, Sugar = 0, Sodium = 350) )
```
This interval means that we are 90% confident that the expected number of calories (true mean of calories) for a food with 24 gram of fats and 0 gram of sugar and 350 mg of sodium is between the calculated lower and upper bounds ("lwr" and "upr" shown above).


**(i)** Use a 90% prediction interval to predict the Calorie content of a Taco Bell Crunchwrap Supreme that has 21g of fat, 6g of sugar, and 1200mg of sodium. Interpret the interval in context.

***Answer***

```{r}
predict(nutrition_model, level = 0.9, interval = 'prediction', newdata = data.frame(Fat = 21, Sugar = 6, Sodium = 1200))
```
This interval means that we are 90% confident that the amount of calories for a food with 21 gram of fats and 6 gram of sugar and 1200 mg of sodium is between the calculated lower and upper bounds ("lwr" and "upr" shown above)..

***

## Exercise 2 (More `lm` for Multiple Regression)

For this exercise we will use the data stored in [`goalies.csv`](goalies.csv). It contains career data for 462 players in the National Hockey League who played goaltender at some point up to and including the 2014-2015 season. The variables in the dataset are:
 
- `W` - Wins
- `GA` - Goals Against
- `SA` - Shots Against
- `SV` - Saves
- `SV_PCT` - Save Percentage
- `GAA` - Goals Against Average
- `SO` - Shutouts
- `MIN` - Minutes
- `PIM` - Penalties in Minutes

For this exercise we will consider three models, each with Wins as the response. The predictors for these models are:

- Model 1: Goals Against, Saves
- Model 2: Goals Against, Saves, Shots Against, Minutes, Shutouts
- Model 3: All Available

**(a)** Use an $F$-test to compares Models 1 and 2. Report the following:

- The null hypothesis
- The value of the test statistic
- The p-value of the test
- A statistical decision at $\alpha = 0.05$
- The model you prefer

***Answer***

```{r}
setwd(".\\")
goalies_data <- read.csv("goalies.csv", header = TRUE, sep = ',')
model1_goalie <- lm(W ~ GA + SV, goalies_data)
model2_goalie <- lm(W ~ GA + SV + SA + MIN + SO , goalies_data)
model3_goalie <- lm(W ~ ., goalies_data)
model1_2_anova <- anova(model1_goalie, model2_goalie)

```

- The null hypothesis

$H_0: \beta_{\texttt{SA}} = \beta_{\texttt{MIN}} = \beta_{\texttt{SO}} = 0$


- The value of the test statistic

```{r}
model1_2_anova$F[2]
```

- The p-value of the test

```{r}
model1_2_anova$`Pr(>F)`[2]
```

- A statistical decision at $\alpha = 0.05$

Reject the null hypotheses

- The model you prefer

Model 2 as its additional predictors are relevant to the response


**(b)** Use an $F$-test to compare Model 3 to your preferred model from part **(a)**. Report the following:

- The null hypothesis
- The value of the test statistic
- The p-value of the test
- A statistical decision at $\alpha = 0.05$
- The model you prefer

***Answer***
```{r}
model2_3_anova <- anova(model2_goalie, model3_goalie)
```

- The null hypothesis

$H_0: \beta_{\texttt{SV_PCT}} = \beta_{\texttt{GAA}} = \beta_{\texttt{PIM}} = 0$

- The value of the test statistic

```{r}
model2_3_anova$F[2]
```

- The p-value of the test

```{r}
model2_3_anova$`Pr(>F)`[2]
```

- A statistical decision at $\alpha = 0.05$

As p-value is less than alpha, we reject the null hypotheses

- The model you prefer
```{r}
summary(model2_goalie)$r.squared
summary(model3_goalie)$r.squared
```

Model 2 is better as it's simpler and its r-squared value is almost close to model 3

**(c)** Use a $t$-test to test $H_0: \beta_{\texttt{SV}} = 0 \ \text{vs} \ H_1: \beta_{\texttt{SV}} \neq 0$ for the model you preferred in part **(b)**. Report the following:

- The value of the test statistic

```{r}
summary(model2_goalie)$coefficients[3,3]
```


- The p-value of the test

```{r}
summary(model2_goalie)$coefficients[3,4]
```

- A statistical decision at $\alpha = 0.05$

Reject the null hypotheses as p-value is less than alpha

***

## Exercise 3 (Regression without `lm`)

For this exercise we will once again use the `Ozone` data from the `mlbench` package. The goal of this exercise is to fit a model with `ozone` as the response and the remaining variables as predictors.

```{r}
data(Ozone, package = "mlbench")
Ozone = Ozone[, c(4, 6, 7, 8)]
colnames(Ozone) = c("ozone", "wind", "humidity", "temp")
Ozone = Ozone[complete.cases(Ozone), ]
```

**(a)** Obtain the estimated regression coefficients **without** the use of `lm()` or any other built-in functions for regression. That is, you should use only matrix operations. Store the results in a vector `beta_hat_no_lm`. To ensure this is a vector, you may need to use `as.vector()`. Return this vector as well as the results of `sum(beta_hat_no_lm ^ 2)`.

***Answer***
```{r}
X <- as.matrix(cbind(rep(1,nrow(Ozone)), Ozone[,2:4]))
Y <- as.matrix(Ozone[,1])
C_Ozone <- solve(t(X) %*% X)
beta_hat_no_lm <- C_Ozone %*% t(X) %*% Y
squared_beta_no_lm <- sum(beta_hat_no_lm ^ 2)
```

**(b)** Obtain the estimated regression coefficients **with** the use of `lm()`. Store the results in a vector `beta_hat_lm`. To ensure this is a vector, you may need to use `as.vector()`. Return this vector as well as the results of `sum(beta_hat_lm ^ 2)`.

***Answer***
```{r}
Ozone_model <- lm(ozone ~ ., Ozone)
beta_hat_lm <- as.vector(Ozone_model$coefficients)
```


**(c)** Use the `all.equal()` function to verify that the results are the same. You may need to remove the names of one of the vectors. The `as.vector()` function will do this as a side effect, or you can directly use `unname()`.

***Answer***
```{r}
all.equal(as.vector(beta_hat_no_lm), beta_hat_lm)
```


**(d)** Calculate $s_e$ without the use of `lm()`. That is, continue with your results from **(a)** and perform additional matrix operations to obtain the result. Output this result. Also, verify that this result is the same as the result obtained from `lm()`.

***Answer***
```{r}
fitted_values <- X %*% beta_hat_no_lm
e_no_lm <- Y - fitted_values
sigma_no_lm <-sqrt((t(e_no_lm) %*% e_no_lm) / (nrow(Y) - nrow(beta_hat_no_lm)))
all.equal(summary(Ozone_model)$sigma, as.vector(sigma_no_lm))
```
Sigma is `r sigma_no_lm`


**(e)** Calculate $R^2$ without the use of `lm()`. That is, continue with your results from **(a)** and **(d)**, and perform additional operations to obtain the result. Output this result. Also, verify that this result is the same as the result obtained from `lm()`.

***Answer***
```{r}
rsquared_no_lm <- 1 - (sum(e_no_lm ^ 2) / sum((Y - mean(Y)) ^ 2))
all.equal(summary(Ozone_model)$r.squared, as.vector(rsquared_no_lm))
```

R-Squared is `r rsquared_no_lm`

***

## Exercise 4 (Regression for Prediction)

For this exercise use the `Auto` dataset from the `ISLR` package. Use `?Auto` to learn about the dataset. The goal of this exercise is to find a model that is useful for **predicting** the response `mpg`. We remove the `name` variable as it is not useful for this analysis. (Also, this is an easier to load version of data from the textbook.)

```{r eval=TRUE}
library(ISLR)
Auto = subset(Auto, select = -c(name))
```

When evaluating a model for prediction, we often look at RMSE. However, if we both fit the model with all the data as well as evaluate RMSE using all the data, we're essentially cheating. We'd like to use RMSE as a measure of how well the model will predict on *unseen* data. If you haven't already noticed, the way we had been using RMSE resulted in RMSE decreasing as models became larger.

To correct for this, we will only use a portion of the data to fit the model, and then we will use leftover data to evaluate the model. We will call these datasets **train** (for fitting) and **test** (for evaluating). The definition of RMSE will stay the same

\[
\text{RMSE}(\text{model, data}) = \sqrt{\frac{1}{n} \sum_{i = 1}^{n}(y_i - \hat{y}_i)^2}
\]

where

- $y_i$ are the actual values of the response for the given data.
- $\hat{y}_i$ are the predicted values using the fitted model and the predictors from the data.

However, we will now evaluate it on both the **train** set and the **test** set separately. So each model you fit will have a **train** RMSE and a **test** RMSE. When calculating **test** RMSE, the predicted values will be found by predicting the response using the **test** data with the model fit using the **train** data. *__Test__ data should never be used to fit a model.*

- Train RMSE: Model fit with *train* data. Evaluate on **train** data.
- Test RMSE: Model fit with *train* data. Evaluate on **test** data.

Set a seed of `1`, and then split the `Auto` data into two datasets, one called `auto_trn` and one called `auto_tst`. The `auto_trn` data frame should contain 292 randomly chosen observations. The `auto_tst` data will contain the remaining observations. Hint: consider the following code:

```{r, eval = TRUE}
set.seed(1)
auto_trn_idx = sample(1:nrow(Auto), 292)
auto_trn <- Auto[auto_trn_idx,]
auto_tst <- Auto[-auto_trn_idx,]
```

Fit a total of five models using the training data.

- One must use all possible predictors.
- One must use only `displacement` as a predictor.
- The remaining three you can pick to be anything you like. One of these should be the *best* of the five for predicting the response.

For each model report the **train** and **test** RMSE. Arrange your results in a well-formatted markdown table. Argue that one of your models is the best for predicting the response.

***Answer***

```{r}
auto_model1 <- lm(mpg ~ ., auto_trn)
auto_model2 <- lm(mpg ~ displacement, auto_trn)
auto_model3 <- lm(mpg ~ weight + year + origin, auto_trn)
auto_model4 <- lm(mpg ~ weight + displacement + acceleration, auto_trn)
auto_model5 <- lm(mpg ~ weight + year + displacement + origin, auto_trn)

auto_models <- list()
auto_models[[1]] <- auto_model1
auto_models[[2]] <- auto_model2
auto_models[[3]] <- auto_model3
auto_models[[4]] <- auto_model4
auto_models[[5]] <- auto_model5


calculate_RMSE = function(fitted_values, y_values){
  return ( sqrt(sum((y_values - fitted_values) ^ 2) / length(y_values)) )
}

auto_rmse_trn <- rep(0,5)
auto_rmse_tst <- rep(0,5)
auto_r_squared <- rep(0,5)

for (i in 1:length(auto_models)){
    auto_rmse_tst[i] <- calculate_RMSE(predict(auto_models[[i]], auto_tst[,2:8]), auto_tst[,1])
    
    auto_rmse_trn[i] <- calculate_RMSE(auto_models[[i]]$fitted.values , auto_trn[,1])
    auto_r_squared[i] <- summary(auto_models[[i]])$r.squared
    
}



library(knitr)
library(mlbench)
df_results <- data.frame(RMSE_Train = rep(0,5), RMSE_Test = rep(0,5))
df_results[,1] <- auto_rmse_trn
df_results[,2] <- auto_rmse_tst

row.names(df_results) = c('All predictors','Displacement','Weight/Year/Origin','Weight/Displacement/Acceleration', 'Weight/Year/Displacement/Origin')
kable(df_results)

```

Based on these results, the last model (predictors: Weight/Year/Displacement/Origin) is the best as it has the least RMSE for the testing set which matters more than training RMSE. And although training RMSE is less for the moel with all predictors, last model is simpler and expected to generalize better. 


***

## Exercise 5 (Simulating Multiple Regression)

For this exercise we will simulate data from the following model:

\[
Y_i = \beta_0 + \beta_1 x_{i1} + \beta_2 x_{i2} + \beta_3 x_{i3} + \beta_4 x_{i4} + \beta_5 x_{i5} + \epsilon_i
\]

Where $\epsilon_i \sim N(0, \sigma^2).$ Also, the parameters are known to be:

- $\beta_0 = 2$
- $\beta_1 = -0.75$
- $\beta_2 = 1.5$
- $\beta_3 = 0$
- $\beta_4 = 0$
- $\beta_5 = 2$
- $\sigma^2 = 25$

We will use samples of size `n = 42`.

We will verify the distribution of $\hat{\beta}_2$ as well as investigate some hypothesis tests.

**(a)** We will first generate the $X$ matrix and data frame that will be used throughout the exercise. Create the following nine variables:

- `x0`: a vector of length `n` that contains all `1`
- `x1`: a vector of length `n` that is randomly drawn from a normal distribution with a mean of `0` and a standard deviation of `2`
- `x2`: a vector of length `n` that is randomly drawn from a uniform distribution between `0` and `4`
- `x3`: a vector of length `n` that is randomly drawn from a normal distribution with a mean of `0` and a standard deviation of `1`
- `x4`: a vector of length `n` that is randomly drawn from a uniform distribution between `-2` and `2`
- `x5`: a vector of length `n` that is randomly drawn from a normal distribution with a mean of `0` and a standard deviation of `2`
- `X`: a matrix that contains `x0`, `x1`, `x2`, `x3`, `x4`, and `x5` as its columns
- `C`: the $C$ matrix that is defined as $(X^\top X)^{-1}$
- `y`: a vector of length `n` that contains all `0`
- `sim_data`: a data frame that stores `y` and the **five** *predictor* variables. `y` is currently a placeholder that we will update during the simulation.

Report the sum of the diagonal of `C` as well as the 5th row of `sim_data`. For this exercise we will use the seed `420`. Generate the above variables in the order listed after running the code below to set a seed.

***Answer***

```{r}
set.seed(420)
sample_size = 42

x0 <- rep(1, sample_size)
x1 <- rnorm(sample_size, 0, 2)
x2 <- runif(sample_size, 0, 4)
x3 <- rnorm(sample_size, 0, 1)
x4 <- runif(sample_size, -2, 2)
x5 <- rnorm(sample_size, 0, 2)

X <- cbind(x0, x1, x2, x3, x4, x5)
C <- solve(t(X) %*% X)
y <- rep(0, sample_size)
sim_data = data.frame(y = y, x0 = x0, x1 = x1, x2 = x2, x3 = x3, x4 = x4, x5 = x5)

sum(diag(C))
sim_data[5,]

```

**(b)** Create three vectors of length `2500` that will store results from the simulation in part **(c)**. Call them `beta_hat_1`, `beta_3_pval`, and `beta_5_pval`.

***Answer***
```{r}
num_simulations <- 2500
beta_hat_1 <- rep(0, num_simulations)
beta_3_pval <- rep(0, num_simulations)
beta_5_pval <- rep(0, num_simulations)
```


**(c)** Simulate 2500 samples of size `n = 42` from the model above. Each time update the `y` value of `sim_data`. Then use `lm()` to fit a multiple regression model. Each time store:

- The value of $\hat{\beta}_1$ in `beta_hat_1`
- The p-value for the two-sided test of $\beta_3 = 0$ in `beta_3_pval`
- The p-value for the two-sided test of $\beta_5 = 0$ in `beta_5_pval`

***Answer***
```{r}

beta_0 = 2
beta_1 = -0.75
beta_2 = 1.5
beta_3 = 0
beta_4 = 0
beta_5 = 2
sigma = 5

for(i in 1:num_simulations)
{
  sim_data$y = beta_0 + (beta_1 * x1) + (beta_2 * x2) + (beta_3 * x3) + (beta_4 * x4) + (beta_5 * x5) + rnorm(sample_size, 0, sigma)
  
  simulated_model <- lm(y ~ ., sim_data)
  beta_hat_1[i] <- coef(summary(simulated_model))[2]
  beta_3_pval[i] <- summary(simulated_model)$coefficients[4,4]
  beta_5_pval[i] <- summary(simulated_model)$coefficients[6,4]
}
```


**(d)** Based on the known values of $X$, what is the true distribution of $\hat{\beta}_1$?

***Answer***
It's a normal distribution with mean equal to -0.75 and variance equal to `r (sigma ^ 2) * C[2,2]` (calculated inline)

**(e)** Calculate the mean and variance of `beta_hat_1`. Are they close to what we would expect? Plot a histogram of `beta_hat_1`. Add a curve for the true distribution of $\hat{\beta}_1$. Does the curve seem to match the histogram?

***Answer***

The mean and variance of beta_hat_1 values are `r mean(beta_hat_1)` and `r sd(beta_hat_1) ^ 2` respectively (calculated inline), they are pretty close to the true mean and variance of beta_1

```{r}
hist(beta_hat_1, prob = TRUE, breaks = 20, xlab = expression(hat(beta)[1]), main = "", border = "dodgerblue")
curve(dnorm(x, mean = beta_1, sd = sqrt((sigma ^ 2) * C[2,2])), 
      col = "darkorange", add = TRUE, lwd = 3)
```

As shown, The true beta curve pretty much matches the histogram.

**(f)** What proportion of the p-values stored in `beta_3_pval` is less than 0.10? Is this what you would expect?

***Answer***
```{r}
sum(beta_3_pval < 0.1) / length(beta_3_pval)
```
When the p-value is less than 0.1, this means we reject the null hypotheses based on 0.1 alpha value, which happened only 10% of the times with the simulations. This is what I would expect for a true beta value of 0 for that prameter as a good model should rarely reject the null hypotheses.


**(g)** What proportion of the p-values stored in `beta_5_pval` is less than 0.01? Is this what you would expect?

***Answer***
```{r}
sum(beta_5_pval < 0.01) / length(beta_5_pval)
```
When the p-value is less than 0.01, this means we reject the null hypotheses for that beta parameter based on 0.01 alpha value, which happened around 80% of the times with the simulations. This is not what I would expect for a true beta value of 0 for that prameter, however, as alpha value is pretty small, we weren't very precise in failing to reject the null hpotheses.
