---
title: "Week 2 - Homework"
author: "STAT 420, Summer 2018, Unger"
date: ''
output:
  html_document: 
    toc: yes
  pdf_document: default
urlcolor: cyan
---

***
***Name*** : Hussein Ahmed Elmessilhy

***NetID***: hae2

## Exercise 1 (Using `lm`)

For this exercise we will use the `cats` dataset from the `MASS` package. You should use `?cats` to learn about the background of this dataset.

**(a)** Suppose we would like to understand the size of a cat's heart based on the body weight of a cat. Fit a simple linear model in `R` that accomplishes this task. Store the results in a variable called `cat_model`. Output the result of calling `summary()` on `cat_model`.
**Answer**: 
```{r}
library(MASS)
cat_model <- lm(Hwt ~ Bwt, cats)
summary(cat_model)
```

**(b)** Output only the estimated regression coefficients. Interpret $\hat{\beta_0}$ and $\beta_1$ in the *context of the problem*. Be aware that only one of those is an estimate.

**Answer**: 
```{r}
cat_model$coefficients 
```
$\hat{\beta_0}$ is the intercept, which represents the y-value of the fitted regression line at x = 0, while $\beta_1$ is the slope of the fitted line, represented as Bwt in the coefficients.


**(c)** Use your model to predict the heart weight of a cat that weights **2.7** kg. Do you feel confident in this prediction? Briefly explain.

**Answer**: 
```{r}
predict(cat_model, newdata = data.frame( Bwt = c(2.7))) 
```
We should be confident in this "interpolated" estimate as it lies in the range of the original values used to build the regression fitted line which was `r min(cats$Bwt)` and `r max(cats$Bwt)` 


**(d)** Use your model to predict the heart weight of a cat that weights **4.4** kg. Do you feel confident in this prediction? Briefly explain.

**Answer**: 
```{r}
predict(cat_model, newdata = data.frame( Bwt = c(4.4))) 
```
I am not very confident in this "extrapolated" estimate as it doesn't lie in the range of the original values used to build the regression fitted line which was `r min(cats$Bwt)` and `r max(cats$Bwt)` 


**(e)** Create a scatterplot of the data and add the fitted regression line. Make sure your plot is well labeled and is somewhat visually appealing.

**Answer**: 
```{r}
plot(Hwt ~ Bwt, cats, 
     main = "Fitted model for cats heart weight against body weight",
     ylab = "Cat heart weight",
     xlab = "Cat body weight",
     col = "orange")
abline(cat_model, col = "blue")
```


**(f)** Report the value of $R^2$ for the model. Do so directly. Do not simply copy and paste the value from the full output in the console after running `summary()` in part **(a)**

**Answer**: 
```{r}
summary(cat_model)$r.squared
```


***

## Exercise 2 (Writing Functions)

This exercise is a continuation of Exercise 1.

**(a)** Write a function called `get_sd_est` that calculates an estimate of $\sigma$ in one of two ways depending on input to the function. The function should take three arguments as input:

- `fitted_vals` - A vector of fitted values from a model
- `actual_vals` - A vector of the true values of the response
- `mle` - A logical (`TRUE` / `FALSE`) variable which defaults to `FALSE`

The function should return a single value:

- $s_e$ if `mle` is set to `FALSE`.
- $\hat{\sigma}$ if `mle` is set to `TRUE`.

**Answer**: 
```{r}
get_sd_est = function(fitted_vals, actual_vals, mle = FALSE) {
  if (mle == FALSE)
  {
    return (sqrt((sum((actual_vals - fitted_vals) ^ 2)) / (length(fitted_vals) - 2)))
  }
  else
  {
    return (sqrt((sum((actual_vals - fitted_vals) ^ 2)) / (length(fitted_vals))))
  }
}
```


**(b)** Run the function `get_sd_est` on the residuals from the model in Exercise 1, with `mle` set to `FALSE`. Explain the resulting estimate in the context of the model.

**Answer**: 
```{r}
get_sd_est(cat_model$fitted.values, cats$Hwt, FALSE)
```
This value is the residual standard error, which represents the standard deviation of the residuals, which means that the estimates of mean heart weight is typically off by that value.


**(c)** Run the function `get_sd_est` on the residuals from the model in Exercise 1, with `mle` set to `TRUE`. Explain the resulting estimate in the context of the model. Note that we are trying to estimate the same parameter as in part **(b)**.

**Answer**: 
```{r}
get_sd_est(cat_model$fitted.values, cats$Hwt, TRUE)
```
This value is another estimate of the standard deviation, but it assumes a maximum likelihood approach rather than the least squares approach used in the previous question estimation.


**(d)** To check your work, output `summary(cat_model)$sigma`. It should match at least one of **(b)** or **(c)**.

**Answer**: 
```{r}
summary(cat_model)$sigma
```

***

## Exercise 3 (Simulating SLR)

Consider the model

\[
Y_i = 5 + -3 x_i + \epsilon_i
\]

with 

\[
\epsilon_i \sim N(\mu = 0, \sigma^2 = 10.24)
\]

where $\beta_0 = 5$ and $\beta_1 = -3$.

This exercise relies heavily on generating random observations. To make this reproducible we will set a seed for the randomization. Alter the following code to make `birthday` store your birthday in the format: `yyyymmdd`. For example, [William Gosset](https://en.wikipedia.org/wiki/William_Sealy_Gosset), better known as *Student*, was born on June 13, 1876, so he would use:

```{r}
birthday = 19850322
set.seed(birthday)
```

**(a)** Use `R` to simulate `n = 25` observations from the above model. For the remainder of this exercise, use the following "known" values of $x$.

**Answer**: 
```{r}
beta_0 <- 5
beta_1 <- -3
x = runif(n = 25, 0, 10)
y = beta_0 + (beta_1 * x) + rnorm(25, 0, sqrt(10.24))
```

You may use [the `sim_slr ` function provided in the text](http://daviddalpiaz.github.io/appliedstats/simple-linear-regression.html#simulating-slr). Store the data frame this function returns in a variable of your choice. Note that this function calls $y$ `response` and $x$ `predictor`.

**(b)** Fit a model to your simulated data. Report the estimated coefficients. Are they close to what you would expect? Briefly explain.

**Answer**: 
```{r}
simulated_model1 <- lm(y ~ x)
coef(simulated_model1)
```
 The resulted coefficients are relatively close to what I expect, which are the actual $\beta_0$ and $\beta_1$ values.


**(c)** Plot the data you simulated in part **(a)**. Add the regression line from part **(b)** as well as the line for the true model. Hint: Keep all plotting commands in the same chunk.


**Answer**: 
```{r}
plot(y ~ x,
     xlab = "Simulated Predictor Variable",
     ylab = "Simulated Response Variable",
     main = "Simulated Regression Data",
     col  = "grey")
abline(simulated_model1, lwd = 3, lty = 1, col = "orange")
abline(beta_0, beta_1, lwd = 3, lty = 2, col = "purple")
legend("topright", c("Estimate", "Truth"), lty = c(1, 2), lwd = 2,
       col = c("orange", "purple"))
```
       
**(d)** Use `R` to repeat the process of simulating `n = 25` observations from the above model $1500$ times. Each time fit a SLR model to the data and store the value of $\hat{\beta_1}$ in a variable called `beta_hat_1`. Some hints:

- Consider a `for` loop.
- Create `beta_hat_1` before writing the `for` loop. Make it a vector of length $1500$ where each element is `0`.
- Inside the body of the `for` loop, simulate new $y$ data each time. Use a variable to temporarily store this data together with the known $x$ data as a data frame.
- After simulating the data, use `lm()` to fit a regression. Use a variable to temporarily store this output.
- Use the `coef()` function and `[]` to extract the correct estimated coefficient.
- Use `beta_hat_1[i]` to store in elements of `beta_hat_1`.
- See the notes on [Distribution of a Sample Mean](http://daviddalpiaz.github.io/appliedstats/introduction-to-r.html#distribution-of-a-sample-mean) for some inspiration.

You can do this differently if you like. Use of these hints is not required.



**Answer**: 
```{r}
beta_hat_1 <- rep(0, 1500)
n <- 25
for(i in 1:1500)
{
  y = beta_0 + (beta_1 * x) + rnorm(n, 0, sqrt(10.24))
  simulated_model1 <- lm(y ~ x)
  beta_hat_1[i] <- coef(simulated_model1)[2]
}
```

**(e)** Report the mean and standard deviation of `beta_hat_1`. Do either of these look familiar?


**Answer**: 
```{r}
mean(beta_hat_1)
sd(beta_hat_1)
```

The resulted mean is familiar as it's very close to the actual $\beta_1$ values, much closer than the $\beta_1$ version generated by one simulation.

**(f)** Plot a histogram of `beta_hat_1`. Comment on the shape of this histogram.

**Answer**: 
```{r}
hist(beta_hat_1)
```

The shape of the histogram is a normal distribution that its mean is around 3.

***

## Exercise 4 (Be a Skeptic)

Consider the model

\[
Y_i = 3 + 0 \cdot x_i + \epsilon_i
\]

with

\[
\epsilon_i \sim N(\mu = 0, \sigma^2 = 4)
\]

where $\beta_0 = 3$ and $\beta_1 = 0$.

Before answering the following parts, set a seed value equal to **your** birthday, as was done in the previous exercise.

```{r}
birthday = 19850322
set.seed(birthday)
beta_0 <- 3
beta_1 <- 0
sigma <- 2
```

**(a)** Use `R` to repeat the process of simulating `n = 75` observations from the above model $2500$ times. For the remainder of this exercise, use the following "known" values of $x$.

```{r}
x = runif(n = 75, 0, 10)
```

Each time fit a SLR model to the data and store the value of $\hat{\beta_1}$ in a variable called `beta_hat_1`. You may use [the `sim_slr ` function provided in the text](http://daviddalpiaz.github.io/appliedstats/simple-linear-regression.html#simulating-slr). Hint: Yes $\beta_1 = 0$.

**Answer**: 
```{r}
beta_hat_1 <- rep(0, 2500)
for(i in 1:2500)
{
  y = beta_0 + (beta_1 * x) + rnorm(length(x), 0, sigma)
  simulated_model1 <- lm(y ~ x)
  beta_hat_1[i] <- coef(simulated_model1)[2]
}
```

**(b)** Plot a histogram of `beta_hat_1`. Comment on the shape of this histogram.

**Answer**: 
```{r}
hist(beta_hat_1)
```

The shape of the histogram is very close to a normal distribution with its mean around to 0.


**(c)** Import the data in [`skeptic.csv`](skeptic.csv) and fit a SLR model. The variable names in `skeptic.csv` follow the same convention as those returned by `sim_slr()`. Extract the fitted coefficient for $\beta_1$.


**Answer**: 
```{r}
setwd(".\\")
skeptic_data <- read.csv(".\\skeptic.csv")
skeptic_model <- lm(response ~ predictor, skeptic_data)
skeptic_beta_1 <- coef(skeptic_model)[2]
```

**(d)** Re-plot the histogram from **(b)**. Now add a vertical red line at the value of $\hat{\beta_1}$ in part **(c)**. To do so, you'll need to use `abline(v = c, col = "red")` where `c` is your value.

**Answer**: 
```{r}
hist(beta_hat_1)
abline(v = skeptic_beta_1, col = "red")
```


**(e)** Your value of $\hat{\beta_1}$ in **(c)** should be negative. What proportion of the `beta_hat_1` values is smaller than your $\hat{\beta_1}$? Return this proportion, as well as this proportion multiplied by `2`.

**Answer**: 
Proportions and double the proportions are :
```{r}
c(sum(beta_hat_1 < skeptic_beta_1) / length(beta_hat_1), sum(beta_hat_1 < skeptic_beta_1) / length(beta_hat_1) * 2)
```

**(f)** Based on your histogram and part **(e)**, do you think the [`skeptic.csv`](skeptic.csv) data could have been generated by the model given above? Briefly explain.

**Answer**: 
It's possible that Skeptic data could have been generated by the model given, however this is possible with a very low probability, and that's because most of the simulated data runs generated a bigger beta_1 value (only 0.0008 proportion and 2 instances had a smaller $\beta_1$ value).

***

## Exercise 5 (Comparing Models)

For this exercise we will use the `Ozone` dataset from the `mlbench` package. You should use `?Ozone` to learn about the background of this dataset. You may need to install the `mlbench` package. If you do so, do not include code to install the package in your `R` Markdown document.

For simplicity, we will perform some data cleaning before proceeding.

```{r eval=TRUE}
library(knitr)
library(mlbench)
data(Ozone, package = "mlbench")
Ozone = Ozone[, c(4, 6, 7, 8)]
colnames(Ozone) = c("ozone", "wind", "humidity", "temp")
Ozone = Ozone[complete.cases(Ozone), ]
```

We have:

- Loaded the data from the package
- Subset the data to relevant variables
    - This is not really necessary (or perhaps a good idea) but it makes the next step easier
- Given variables useful names
- Removed any observation with missing values
    - This should be given much more thought in practice

For this exercise we will define the "Root Mean Square Error" of a model as

\[
\text{RMSE} = \sqrt{\frac{1}{n} \sum_{i = 1}^{n}(y_i - \hat{y}_i)^2}.
\]

**(a)** Fit three SLR models, each with "ozone" as the response. For the predictor, use "wind speed," "humidity percentage," and "temperature" respectively. For each, calculate $\text{RMSE}$ and $R^2$. Arrange the results in a markdown table, with a row for each model. Suggestion: Create a data frame that stores the results, then investigate the `kable()` function from the `knitr` package.

***Answer***
```{r eval=TRUE}
df_results <- data.frame(R.Squared = rep(0,3), RMSE = rep(0,3))
row.names(df_results) <- c("Wind model", "Humidity model", "Temp. model")
wind_model <- lm(ozone~wind, Ozone) 
humidity_model <- lm(ozone~humidity, Ozone) 
temp_model <- lm(ozone~temp, Ozone) 

RMSE = function(y_values, fitted_values){
  return ( sqrt(sum( (y_values - fitted_values)^2) / length(y_values) ) )
}

df_results[1,] <- c(summary(wind_model)$r.squared, RMSE(Ozone$ozone, wind_model$fitted.values))
df_results[2,] <- c(summary(humidity_model)$r.squared, RMSE(Ozone$ozone, humidity_model$fitted.values))
df_results[3,] <- c(summary(temp_model)$r.squared, RMSE(Ozone$ozone, temp_model$fitted.values))
kable(df_results)

```


**(b)** Based on the results, which of the three predictors used is most helpful for predicting ozone readings? Briefly explain.

***Answer***:
The temperature model is no doubt the most helpful for predicting the ozone readings, this is justified by the relatively high R-Squared value as well as the relatively smaller root mean squared error. 

***
