---
title: "Week 3 - Homework"
author: "STAT 420, Summer 2018, Unger"
date: ''
output:
  html_document: 
    toc: yes
  pdf_document: default
urlcolor: cyan
---

***

## Exercise 1 (Using `lm` for Inference)

For this exercise we will use the `cats` dataset from the `MASS` package. You should use `?cats` to learn about the background of this dataset.

**(a)** Fit the following simple linear regression model in `R`. Use heart weight as the response and body weight as the predictor. 

\[
Y_i = \beta_0 + \beta_1 x_i + \epsilon_i
\]

Store the results in a variable called `cat_model`. Use a $t$ test to test the significance of the regression. Report the following:

- The null and alternative hypotheses
- The value of the test statistic
- The p-value of the test
- A statistical decision at $\alpha = 0.05$
- A conclusion in the context of the problem

When reporting these, you should explicitly state them in your document, not assume that a reader will find and interpret them from a large block of `R` output.

***Answer***
```{r}
library(MASS)
cat_model <- lm(Hwt~Bwt, cats)
summary(cat_model)
```
- The null and alternative hypotheses
  Null hypotheses is that $\beta_1 = 0$ and alternative hypotheses is that $\beta_1 \neq  0$ 
- The value of the test statistic
  -0.515 for $\beta_0$ and 16.119 for $\beta_1$
- The p-value of the test
  0.607 for $\beta_0$ and 2e-16 for $\beta_1$
- A statistical decision at $\alpha = 0.05$
  Reject the null hypotheses
- A conclusion in the context of the problem
  As we failed to reject the null hypotheses, there's a linear relationship between   the body weight and the heart weight

**(b)** Calculate a 90% confidence interval for $\beta_1$. Give an interpretation of the interval in the context of the problem.

***Answer***
```{r}
confint(cat_model, level = 0.9)[2,]
```

**(c)** Calculate a 99% confidence interval for $\beta_0$. Give an interpretation of the interval in the context of the problem.

***Answer***
```{r}
confint(cat_model, level = 0.99)[1,]
```

**(d)** Use a 99% confidence interval to estimate the mean heart weight for body weights of 2.1 and 2.8 kilograms. Which of the two intervals is wider? Why?

***Answer***
```{r}
predict(cat_model, level = 0.99, newdata = data.frame(Bwt = c(2.1, 2.8)), interval = c('confidence'))

c(min(cats$Bwt), max(cats$Bwt))
mean(cats$Bwt)
```

**(e)** Use a 99% prediction interval to predict the heart weight for body weights of 2.8 and 4.2 kilograms.

***Answer***
```{r}
predict(cat_model, level = 0.99, newdata = data.frame(Bwt = c(2.8, 4.2)), interval = c('prediction'))
```

**(f)** Create a scatterplot of the data. Add the regression line, 90% confidence bands, and 90% prediction bands.

***Answer***
```{r}
Bwt_grid = seq(min(cats$Bwt), max(cats$Bwt), by = 0.01)
dist_ci_band = predict(cat_model, 
                       newdata = data.frame(Bwt = Bwt_grid), 
                       interval = "confidence", level = 0.9)
dist_pi_band = predict(cat_model, 
                       newdata = data.frame(Bwt = Bwt_grid), 
                       interval = "prediction", level = 0.9) 

plot(Hwt ~ Bwt, data = cats,
     xlab = "Cat body weight",
     ylab = "Cat heart weight",
     main = "Cat heart weight vs body weight",
     pch  = 20,
     cex  = 1,
     col  = "grey",
     ylim = c(min(dist_pi_band), max(dist_pi_band)))
abline(cat_model, lwd = 3, col = "darkorange")

lines(Bwt_grid, dist_ci_band[,"lwr"], col = "dodgerblue", lwd = 3, lty = 2)
lines(Bwt_grid, dist_ci_band[,"upr"], col = "dodgerblue", lwd = 3, lty = 2)
lines(Bwt_grid, dist_pi_band[,"lwr"], col = "dodgerblue", lwd = 3, lty = 3)
lines(Bwt_grid, dist_pi_band[,"upr"], col = "dodgerblue", lwd = 3, lty = 3)
points(mean(cats$Bwt), mean(cats$Hwt), pch = "+", cex = 3)
```

**(g)** Use a $t$ test to test:

- $H_0: \beta_1 = 4$
- $H_1: \beta_1 \neq 4$

Report the following:

- The value of the test statistic
- The p-value of the test
- A statistical decision at $\alpha = 0.05$

When reporting these, you should explicitly state them in your document, not assume that a reader will find and interpret them from a large block of `R` output.

***Answer***
```{r}
t_val_4 <- (summary(cat_model)$coefficients[2,1] - 4) / summary(cat_model)$coefficients[2,2]

p_val_4 <- 2 * pt(t_val_4,  df = length(resid(cat_model)) - 2, lower.tail = FALSE )

#fail to reject the null hypotheses as p val is high
```

***

## Exercise 2 (More `lm` for Inference)

For this exercise we will use the `Ozone` dataset from the `mlbench` package. You should use `?Ozone` to learn about the background of this dataset. You may need to install the `mlbench` package. If you do so, do not include code to install the package in your `R` Markdown document.

For simplicity, we will re-perform the data cleaning done in the previous homework.

```{r}
data(Ozone, package = "mlbench")
Ozone = Ozone[, c(4, 6, 7, 8)]
colnames(Ozone) = c("ozone", "wind", "humidity", "temp")
Ozone = Ozone[complete.cases(Ozone), ]
```

**(a)** Fit the following simple linear regression model in `R`. Use the ozone measurement as the response and wind speed as the predictor. 

\[
Y_i = \beta_0 + \beta_1 x_i + \epsilon_i
\]

Store the results in a variable called `ozone_wind_model`. Use a $t$ test to test the significance of the regression. Report the following:

- The null and alternative hypotheses
- The value of the test statistic
- The p-value of the test
- A statistical decision at $\alpha = 0.01$
- A conclusion in the context of the problem

When reporting these, you should explicitly state them in your document, not assume that a reader will find and interpret them from a large block of `R` output.

***Answer***
```{r}
ozone_wind_model <- summary(lm(ozone ~ wind, Ozone))
```
- The null and alternative hypotheses
  $H_0: \beta_1 = 0$
  $H_1: \beta_1 \neq 0$
  
- The value of the test statistic
```{r}
ozone_wind_model$coefficients[2,3]
```

- The p-value of the test
```{r}
ozone_wind_model$coefficients[2,4]
```

- A statistical decision at $\alpha = 0.01$
We fail to reject the null hypotheses as p-value < $\alpha$ is `r ozone_wind_model$coefficients[2,4] < 0.01`

- A conclusion in the context of the problem
Conclusion is that there are no linear relationship between the predictor (wind variable) and the response (ozone variable)

**(b)** Fit the following simple linear regression model in `R`. Use the ozone measurement as the response and temperature as the predictor. 

\[
Y_i = \beta_0 + \beta_1 x_i + \epsilon_i
\]

Store the results in a variable called `ozone_temp_model`. Use a $t$ test to test the significance of the regression. Report the following:

- The null and alternative hypotheses
- The value of the test statistic
- The p-value of the test
- A statistical decision at $\alpha = 0.01$
- A conclusion in the context of the problem

When reporting these, you should explicitly state them in your document, not assume that a reader will find and interpret them from a large block of `R` output.


***Answer***
```{r}
ozone_temp_model <- summary(lm(ozone ~ temp, Ozone))
```
- The null and alternative hypotheses
  $H_0: \beta_1 = 0$
  $H_1: \beta_1 \neq 0$
  
- The value of the test statistic
```{r}
ozone_temp_model$coefficients[2,3]
```

- The p-value of the test
```{r}
ozone_temp_model$coefficients[2,4]
```

- A statistical decision at $\alpha = 0.01$
We fail to reject the null hypotheses as p-value < $\alpha$ is `r ozone_temp_model$coefficients[2,4] < 0.01`

- A conclusion in the context of the problem
Conclusion is that there are no linear relationship between the predictor (wind variable) and the response (ozone variable)

***

## Exercise 3 (Simulating Sampling Distributions)

For this exercise we will simulate data from the following model:

\[
Y_i = \beta_0 + \beta_1 x_i + \epsilon_i
\]

Where $\epsilon_i \sim N(0, \sigma^2).$ Also, the parameters are known to be:

- $\beta_0 = -5$
- $\beta_1 = 3.25$
- $\sigma^2 = 16$

We will use samples of size $n = 50$.

**(a)** Simulate this model $2000$ times. Each time use `lm()` to fit a simple linear regression model, then store the value of $\hat{\beta}_0$ and $\hat{\beta}_1$. Set a seed using **your** birthday before performing the simulation. Note, we are simulating the $x$ values once, and then they remain fixed for the remainder of the exercise.

```{r}
birthday = 19850323
set.seed(birthday)
n = 50
x = seq(0, 10, length = n)


beta_0 <- -5
beta_1 <- 3.25
sigma <- 4

beta_hat_1 <- rep(0, 2000)
beta_hat_0 <- rep(0, 2000)
n <- 25
for(i in 1:length(beta_hat_1))
{
  y = beta_0 + (beta_1 * x) + rnorm(n, 0, sigma)
  simulated_model <- lm(y ~ x)
  beta_hat_0[i] <- coef(simulated_model)[1]
  beta_hat_1[i] <- coef(simulated_model)[2]
}

```

**(b)** Create a table that summarizes the results of the simulations. The table should have two columns, one for $\hat{\beta}_0$ and one for $\hat{\beta}_1$. The table should have four rows:

- A row for the true expected value given the known values of $x$
- A row for the mean of the simulated values
- A row for the true standard deviation given the known values of $x$
- A row for the standard deviation of the simulated values

***Answer***

```{r}
library(knitr)
library(mlbench)
df_results <- data.frame(beta_0_hat = rep(0,4), beta_0_hat = rep(0,4))
df_results[1,1] <- beta_0
df_results[1,2] <- beta_1

df_results[2,1] <- mean(beta_hat_0)
df_results[2,2] <- mean(beta_hat_1)

Sxx = sum((x - mean(x)) ^ 2)
var_beta_1_hat = sigma ^ 2 / Sxx
var_beta_0_hat = sigma ^ 2 * (1 / n + mean(x) ^ 2 / Sxx)

df_results[3,1] <- sqrt(var_beta_0_hat)
df_results[3,2] <- sqrt(var_beta_1_hat)

df_results[4,1] <- sd(beta_hat_0)
df_results[4,2] <- sd(beta_hat_1)

row.names(df_results) = c('True Expected Value','Simulated mean','True standard deviation','Simulated standard deviation')
kable(df_results)
```


**(c)** Plot two histograms side-by-side:

- A histogram of your simulated values for $\hat{\beta}_0$. Add the normal curve for the true sampling distribution of $\hat{\beta}_0$.
- A histogram of your simulated values for $\hat{\beta}_1$. Add the normal curve for the true sampling distribution of $\hat{\beta}_1$.

***Answer***
```{r}
par(mfrow=c(1,2))
hist(beta_hat_0, prob = TRUE, breaks = 20, xlab = expression(hat(beta)[0]), main = "", border = "dodgerblue")
curve(dnorm(x, mean = beta_0, sd = sqrt(var_beta_0_hat)), 
      col = "darkorange", add = TRUE, lwd = 3)

hist(beta_hat_1, prob = TRUE, breaks = 20, xlab = expression(hat(beta)[1]), main = "", border = "dodgerblue")
curve(dnorm(x, mean = beta_1, sd = sqrt(var_beta_1_hat)), 
      col = "darkorange", add = TRUE, lwd = 3)
```

***

## Exercise 4 (Simulating Confidence Intervals)

For this exercise we will simulate data from the following model:

\[
Y_i = \beta_0 + \beta_1 x_i + \epsilon_i
\]

Where $\epsilon_i \sim N(0, \sigma^2).$ Also, the parameters are known to be:

- $\beta_0 = 5$
- $\beta_1 = 2$
- $\sigma^2 = 9$

We will use samples of size $n = 25$.

Our goal here is to use simulation to verify that the confidence intervals really do have their stated confidence level. Do **not** use the `confint()` function for this entire exercise.

**(a)** Simulate this model $2500$ times. Each time use `lm()` to fit a simple linear regression model, then store the value of $\hat{\beta}_1$ and $s_e$. Set a seed using **your** birthday before performing the simulation. Note, we are simulating the $x$ values once, and then they remain fixed for the remainder of the exercise.

```{r}
birthday = 18760613
set.seed(birthday)
n = 25
x = seq(0, 2.5, length = n)
```

**(b)** For each of the $\hat{\beta}_1$ that you simulated, calculate a 95% confidence interval. Store the lower limits in a vector `lower_95` and the upper limits in a vector `upper_95`. Some hints:

- You will need to use `qt()` to calculate the critical value, which will be the same for each interval.
- Remember that `x` is fixed, so $S_{xx}$ will be the same for each interval.
- You could, but do not need to write a `for` loop. Remember vectorized operations.

**(c)** What proportion of these intervals contains the true value of $\beta_1$?

**(d)** Based on these intervals, what proportion of the simulations would reject the test $H_0: \beta_1 = 0$ vs $H_1: \beta_1 \neq 0$ at $\alpha = 0.05$?

**(e)** For each of the $\hat{\beta}_1$ that you simulated, calculate a 99% confidence interval. Store the lower limits in a vector `lower_99` and the upper limits in a vector `upper_99`.

**(f)** What proportion of these intervals contains the true value of $\beta_1$?

**(g)** Based on these intervals, what proportion of the simulations would reject the test $H_0: \beta_1 = 0$ vs $H_1: \beta_1 \neq 0$ at $\alpha = 0.01$?

***

## Exercise 5 (Prediction Intervals "without" `predict`)

Write a function named `calc_pred_int` that performs calculates prediction intervals:

$$
\hat{y}(x) \pm t_{\alpha/2, n - 2} \cdot s_e\sqrt{1 + \frac{1}{n}+\frac{(x-\bar{x})^2}{S_{xx}}}.
$$

for the linear model

$$
Y_i = \beta_0 + \beta_1 x_i + \epsilon_i.
$$

**(a)** Write this function. You may use the `predict()` function, but you may **not** supply a value for the `level` argument of `predict()`. (You can certainly use `predict()` any way you would like in order to check your work.)

The function should take three inputs:

- `model`, a model object that is the result of fitting the SLR model with `lm()`
- `newdata`, a data frame with a single observation (row)
    - This data frame will need to have a variable (column) with the same name as the data used to fit `model`.
- `level`, the level (0.90, 0.95, etc) for the interval with a default value of `0.95`

The function should return a named vector with three elements:

- `estimate`, the midpoint of the interval
- `lower`, the lower bound of the interval
- `upper`, the upper bound of the interval

**(b)** After writing the function, run this code:

```{r, eval = FALSE}
newcat_1 = data.frame(Bwt = 4.0)
calc_pred_int(cat_model, newcat_1)
```

**(c)** After writing the function, run this code:

```{r, eval = FALSE}
newcat_2 = data.frame(Bwt = 3.3)
calc_pred_int(cat_model, newcat_2, level = 0.99)
```


